Implementing simplified INLA into `aghq`

### 2022-11-17 src/dev_sinla/sinla.Rmd
Change the notation to match current paper draft, and fix a few mistakes I noticed

The AGHQ/ELGM approximation to the joint posterior of the latent field is the following object
$$
\begin{equation}
\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega_k(z), (\#eq:pxgiveny)
\end{equation}
$$
where $\theta(z) \in \Theta$ and $\omega_k(z) \in \mathbb{R}$ are AGHQ nodes and weights generated by "adaption" to $\tilde p_\text{LA}(\theta \, | \, y)$.
In particular, let
$$
\mathcal{Q}(1, k) = \{z_d \in \mathbb{R}: H_k(z_d) = 0\}
$$
be the set of nodes in one-dimension, exactly the zeros of the Hermite polynomials $H_k(z_d)$ for $k \in \mathbb{N}$, and
$$
\omega_k(z_d) = \frac{k!}{H_{k + 1}(z_d)^2 \times \phi(z_d)}, 
$$
be the associated weights.
Define $\mathcal{Q}(d, k) = \mathcal{Q}(1, k)^d$ to be the set of nodes $z$ in $d$-dimensions, of which there will be $k^d$.
Then these points are adapted via $\theta(z) = \hat \theta + Lz$, where $\hat \theta$ is the mode (of $p_\text{LA}(\theta \, | \, y)$) and $L$ is the lower Cholesky triangle of the curvature (at the mode) such that $LL^\top = H^{-1}$.

This can be evaluated at any point, though in practice we work with it by drawing samples $x \sim p_\text{G}(x \, | \, \theta, y)$ using a method from sampling from a Gaussian given its mean and precision from Rue (in particular you solve a linear equation using the lower Cholesky triangle of the curvature at a randomly chosen quadrature point).
For the $i$th marginal, we just take the appropriate samples from the joint.
The approximation is therefore
$$
\begin{equation}
\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega_k(z), (\#eq:pxigiveny)
\end{equation}
$$
where $p_\text{G}(x_i \, | \, \theta, y)$ just takes the $i$th elements of the mean vector and $(i, i)$th element of the covariance matrix.

@@

The goal of this notebook ([Implementing simplified INLA into `aghq`](https://athowes.github.io/elgm-inf/sinla.html)) is to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation $p_{\text{SINLA}}(x_i \, | \, \theta, y)$ as follows
$$
\tilde p(x_i \, | \, \theta, y) = \sum_z \tilde p_{\text{SINLA}}(x_i \, | \, \theta, y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega_k(z).
$$
We want to do this because the new approximation should be more accurate than taking marginals of a joint Gaussian approximation.
At the extreme, you could just compute a Laplace approximation for each marginal, integrating out all other elements of the latent field, $x_{-i}$, together with the fixed effects $\theta$.
This would be an accurate approach, but not computationally feasible for large models.
What we hope to achieve is an approximation which gets to most of the accuracy benefits of the Laplace approximation approach, but without the high computational cost.

@@

The Gaussian approximation $p_\text{G}(x \, | \, \theta, y)$ is previously used in computing the Laplace approximation to $p(\theta \, | \, y)$
$$
\begin{equation}
\tilde p(\theta \, | \, y) \propto \frac{p(x, \theta, y)}{p_\text{G}(x \, | \, \theta, y)} \Big\rvert_{x = \mu(\theta)} = \frac{p(y, \mu(\theta), \theta)}{\det \left(\frac{1}{2 \pi} Q(\theta) \right)^{1/2}}, (\#eq:pthetagiveny)
\end{equation}
$$
such that one easy way to calculate it is to set `random = "x"` in the `TMB` template to integrate the latent field out, then recover the Gaussian approximation from there, as we have done in the function `h` above.

Let's optimise `h` to obtain estimates $\hat \theta$ which maximise $\tilde p(\theta \, | \, y)$:

@@

We can also look at the parameters -- that is, mean vector and precision matrix -- of the Gaussian approximation at the mode `opt_theta$par` above.
Note that in `aghq`, what we would do is create a quadrature grid, then get the mean vector and precision matrix for each value of the quadrature grid over $\theta$.
By just sticking to the mode of $\theta$, we are effectively doing `aghq` with $k = 1$ in each dimension, i.e. one grid point, corresponding to empirical Bayes type of inference.

@@

#' Note that I'm calling this mean, but because it's a Gaussian the mean is the same as the mode
x_mean <- mm[h$env$random]
x_mean <- rlang::duplicate(x_mean) #' Insurance...
length(x_mean) #' The length is 30

#' Hessian evaluated at the mean (or equivalently mode)
x_Q <- h$env$spHess(mm, random = TRUE)
x_Q <- rlang::duplicate(x_Q)
dim(x_Q) #' With dimension [30 x 30]
```

@@

Given $\theta$, we compute the `mean` and `Q` of $x$ (corresponding to `phi_prev`) using `h`.
We then concatenate and evaluate using `f` to get the numerator, multiplying by -1 because `f$fn` gives the negative log-likelihood, and we want the log-likelihood.
In the denominator, we use `determinant` on the log-scale as it is more numerically stable than e.g. `log(det(...))`.

```{r}
calculate_rhs <- function(theta) {
  #' Calculate mean and Q for those theta
  #' TMB template remembers what you've called -- spooky
  h$fn(theta)
  mm <- h$env$last.par
  mean <- mm[h$env$random]
  Q <- h$env$spHess(mm, random = TRUE)

  #' Append the calculated latent field mean
  param <- theta
  param$phi_prev <- as.numeric(mean)
  param <- param[c("beta_prev", "phi_prev", "log_sigma_phi_prev")]

@@

  #' Evaluate the log-denominator of the RHS
  denom <- 0.5 * (determinant(Q, logarithm = TRUE)$modulus - (nrow(Q) * log(2 * pi)))

@@

```{r}
i <- 1
mean_i <- as.numeric(x_mean[i])
var_i <- solve(x_Q)[i, i]
```

@@

```{r message=FALSE, warning=FALSE}
plot_pdf2 <- plot_pdf +
  stat_function(
    data = data.frame(x = c(0.5, 3.5)), aes(x, col = "Joint Gaussian marginal"), fun = dnorm, n = 101, 
    args = list(mean = mean_i, sd = sqrt(var_i))
  ) +
  labs(x = "prev_phi[1]", y = "Posterior", col = "Approximation") +
  scale_colour_manual(values = cbpalette) +
  theme_minimal()

M <- 1000
samples_gaussian <- rnorm(M, mean_i, sqrt(var_i))

@@

For the approximation to the latent field marginal posteriors, another approach we could take is to calculate the full Laplace approximation
$$
\begin{align*}
  \tilde p(x_i \, | \, \theta, y) &\propto \frac{p(x, \theta, y)}{p_\text{G}(x_{-i} \, | \, x_i, \theta, y)} \Big\rvert_{x_{-i} = \mu_{-i}(x_i, \theta)} (\#eq:pxigiventhetay) \\
  &= \frac{p(y, x_i, \mu_{-i}(x_i, \theta), \theta)}{\det(\frac{1}{2 \pi} Q_{-i}(x_i, \theta))^{1/2}}
\end{align*}
$$
integrating out $x_{-i}$ with a Gaussian approximation.
This is not practical in some settings as it involves recomputing the Gaussian approximation $p_\text{G}(x_{-i} \, | \, x_i, \theta, y)$ for each value of $(x_i, \theta)$ which can be too computationally expensive.
We will still go through the steps for implementing it here though for demonstration.
With of a bit of effort, it can be implemented in `TMB` by passing in data where elements $i$ and $-i$ of the latent field are explicitly named (see Appendix \@ref(appendix) for the `TMB` code):

@@

The mean and precision matrix for the Gaussian approximation $\tilde p_\text{G}(x_{-i} \, | \, x_i, \theta, y)$ are:

```{r}
get_mean_and_Q <- function(obj) {
  mm <- obj$env$last.par
  mean <- mm[obj$env$random]
  mean <- rlang::duplicate(mean)
  Q <- obj$env$spHess(mm, random = TRUE)
  Q <- rlang::duplicate(Q)
  return(list(mean = mean, Q = Q))
}

xminusi <- get_mean_and_Q(obj)
```

@@

```{r}
x_Sigma <- solve(x_Q)

gaussian_df <- lapply(1:dat$n, function(i) {
  x <- seq(-5, 5, length.out = 1000)
  mean_i <- as.numeric(x_mean[i])
  sd_i <- as.numeric(sqrt(x_Sigma[i, i]))
  data.frame(index = i, x = x, pdf = dnorm(x, mean = mean_i, sd = sd_i))
}) %>%
  bind_rows()

samples_gaussian <- lapply(1:dat$n, function(i) {
  mean_i <- as.numeric(x_mean[i])
  sd_i <- as.numeric(sqrt(x_Sigma[i, i]))
  samples <- rnorm(M, mean_i, sd_i)
  data.frame("value" = samples)
}) %>%
  bind_rows(.id = "index")
```

@@

```{r}
ks_gaussian_tmbstan <- lapply(1:length(mean), function(i) {
  samples_gaussian_i <- filter(samples_gaussian, index == i)$value
  samples_tmbstan_i <- filter(samples_tmbstan, index == i)$value
  inf.utils::ks_test(samples_gaussian_i, samples_tmbstan_i)
})

ks_laplace_tmbstan <- lapply(1:length(mean), function(i) {
  samples_laplace_i <- filter(samples_laplace, index == i)$value
  samples_tmbstan_i <- filter(samples_tmbstan, index == i)$value
  inf.utils::ks_test(samples_laplace_i, samples_tmbstan_i)
})

@@

Now we want to evaluate
$$
\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega_k(z).
$$

@@

The weight $\omega_k(z)$ is:

@@

```{r}
mean_i <- as.numeric(modesandhessians$mode[[z]][i])
Sigma <- solve(modesandhessians$H[[z]])
sd_i <- as.numeric(sqrt(Sigma[i, i]))

dnorm(x_test_value, mean = mean_i, sd = sd_i, log = TRUE)
```

@@

```{r}
dnorm(x_test_value, mean = mean_i, sd = sd_i, log = TRUE) + (- h_norm(theta)) + log(modesandhessians$weights[z])
```

To obtain the unlogged version we can use:
$$
\exp \left( \log\tilde p_\text{G}(x_i \, | \, \theta(z), y) + \log \tilde p_\text{LA}(\theta(z) \, | \, y) + \log \omega_k(z) \right)
$$

```{r}
exp(dnorm(x_test_value, mean = mean_i, sd = sd_i, log = TRUE) + (- h_norm(theta)) + log(modesandhessians$weights[z]))
```

@@

for(z in 1:length(modesandhessians$weights)) {
  theta <- as.numeric(modesandhessians[z, thetanames])
  mean_i <- as.numeric(modesandhessians$mode[[z]][i])
  Sigma <- solve(modesandhessians$H[[z]])
  sd_i <- as.numeric(sqrt(Sigma[i, i]))
  pdf <- pdf + exp(dnorm(x_test_value, mean = mean_i, sd = sd_i, log = TRUE) + (- h_norm(theta)) + log(modesandhessians$weights[z]))
}

@@

Let's do as we have above, but with the full Laplace approximation in place of the Gaussian.
We want to evaluate
$$
\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{LA}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega_k(z),
$$
so much of what we computed above in Section \@ref(aghq-gaussian-marginal) is still relevant.
The only thing that has changed is that we're now using $\tilde p_\text{LA}(x_i \, | \, \theta(z), y)$ rather than $\tilde p_\text{G}(x_i \, | \, \theta(z), y)$.

@@

As before, the weight $\omega_k(z)$ is:

@@

<!-- In Section \@ref(full-laplace) we directly calculate the Laplace approximation to $\int p(x, y, \theta) \text{d} \theta_{-i}$. -->
<!-- Suppose that -->
<!-- $$ -->
<!-- x \sim \mathcal{N}(\mu(\theta), Q(\theta)^{-1}) -->
<!-- $$ -->
<!-- then by the conditional distributions of a multivariate Gaussian distribution $x_{-i} \, | \, x_i$ is Gaussian with mean -->
<!-- $$ -->
<!-- \mu(\theta)_{-i} + \Sigma(\theta)_{-i, i} \Sigma(\theta)^{-1}_{i, i} (x_i - \mu (\theta)_i) (\#eq:conditionalmean) -->
<!-- $$ -->
<!-- and precision matrix -->
<!-- $$ -->
<!-- Q(\theta)_{-i, -i} (\#eq:conditionalprecision) -->
<!-- $$ -->

<!-- We follow Wood (2019) Section 2. -->
<!-- In this section we omit dependence on $\theta$ such that: $\mu = \mu(\theta)$, $Q = Q(\theta)$, $\mu(x_i) = \mu(x_i, \theta)$ and $Q(x_i) = Q(x_i, \theta)$. -->

@@

<!-- The first task is to "compute the Cholesky factor of $Q_0 := Q_{-i, -i}$ by update of the Cholesky factor of $Q$". -->

@@

<!-- mm <- h$env$last.par -->
<!-- mean <- mm[h$env$random] -->
<!-- Q <- h$env$spHess(mm, random = TRUE) -->
<!-- Q0 <- Q[-i, -i]  -->

<!-- L0 <- chol(Q0) -->
<!-- ``` -->

<!-- We have that $Q_0 = L_0 {L_0}^\top$: -->

<!-- ```{r} -->
<!-- sum(abs((L0 %*% t(L0)) - Q0)) < 10E-9 -->
<!-- ``` -->

@@

<!-- Use Newton's method with fixed Hessian $Q_{-i, -i}$ to find $\mu(x_i)$ starting from \@ref(eq:conditionalmean). -->
<!-- We start by computing the conditional mean for the first grid point: -->

<!-- ```{r} -->
<!-- Sigma <- solve(Q) -->
<!-- conditional_mean <- mean[-i] + Sigma[-i, i] * Sigma[i, i]^{-1} * (x_grid[1] - mean[i]) -->
<!-- ``` -->

<!-- Note that for this model the latent field is IID, so the conditional mean is the same as the mean. -->
<!-- Now we use Newton's method to find $\mu(x_i)$. -->
<!-- This requires some function $f(\mu(x_i), \ldots) = 0$. -->

@@

<!-- For $j = 1, \ldots, J - 1$ compute -->
<!-- $$ -->
<!-- q = Q_0 \Delta_j + u_j \mathcal{D}_{2j}u_j^\top \Delta_j -->
<!-- $$ -->
<!-- and -->
<!-- $$ -->
<!-- g = \Delta_x \log p(\mu(x_i) + \Delta_j^i, y, \theta). -->
<!-- $$ -->
<!-- Then compute the matrix -->
<!-- $$ -->
<!-- u_{j + 1} = \{q(\Delta_j^\top j)^{-1/2}, g_{-i}(\Delta_j^\top g_{-i})^{-1/2}, u_j\}. -->
<!-- $$ -->

@@

<!-- Compute the determinant approximation -->
<!-- $$ -->
<!-- \det(Q_1) = \det(Q_0) \det(I_{2J} + u_J^\top Q_0^{-1} u_J \mathcal{D}_{2j}). -->
<!-- $$ -->

@@

| Description | Here          | Wood (2019) notation | Notes    |
|:------------|:--------------|:---------------------|:---------|
| Random effects | $x$        | $\beta$              | Stringer (2021) use $w$
| Random effects which maximise (the / some) objective for given $\theta$ | $\mu (\theta)$ | $\hat \beta$ | Wood omits dependence on $\theta$ |
| Random effects which maximise (the / some) objective for given $\theta$, subject to $x_i$ being fixed | $(x_i, \mu(x_i, \theta))$ | $\tilde \beta$ | We have $\mu(x_i, \theta))$ having dimension $n - 1$ |
| Hessian / precision matrix of the negative log-likelihood with respect to $x$ at $\mu(\theta)$ | $Q(\theta)$ | $H$ | Unsure if it's preferable to use Hessian or precision notation here |

>>>>

The AGHQ/ELGM approximation to the joint posterior of the latent field is the following object
$$
\begin{equation}
\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega(z), (\#eq:pxgiveny)
\end{equation}
$$
where $\theta(z) \in \Theta$ and $\omega(z) \in \mathbb{R}$ are AGHQ nodes and weights generated by "adaption" to $\tilde p_\text{LA}(\theta \, | \, y)$.
In particular, let
$$
\mathcal{Q}(1, k) = \{z_d \in \mathbb{R}: H_k(z_d) = 0\}
$$
be the set of nodes in one-dimension, exactly the zeros of the Hermite polynomials $H_k(z_d)$ for $k \in \mathbb{N}$, and
$$
\omega(z_d) = \frac{k!}{H_{k + 1}(z_d)^2 \times \phi(z_d)}, 
$$
be the associated weights.
Define $\mathcal{Q}(d, k) = \mathcal{Q}(1, k)^d$ to be the set of nodes $z$ in $d$-dimensions, of which there will be $k^d$.
Then these points are adapted via $\theta(z) = \hat \theta + Lz$, where $\hat \theta$ is the mode (of $p_\text{LA}(\theta \, | \, y)$) and $L$ is the lower Cholesky triangle of the curvature (at the mode) such that $LL^\top = H^{-1}$.

This can be evaluated at any point, though in practice we work with it by drawing samples $x \sim p_\text{G}(x \, | \, \theta, y)$ using a method from sampling from a Gaussian given its mode and precision from Rue (in particular you solve a linear equation using the lower Cholesky triangle of the curvature at a randomly chosen quadrature point).
For the $i$th marginal, we just take the appropriate samples from the joint.
The approximation is therefore
$$
\begin{equation}
\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega(z), (\#eq:pxigiveny)
\end{equation}
$$
where $p_\text{G}(x_i \, | \, \theta, y)$ just takes the $i$th elements of the mode vector and $(i, i)$th element of the covariance matrix.

@@

The goal of this notebook ([Implementing simplified INLA into `aghq`](https://athowes.github.io/elgm-inf/sinla.html)) is to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation $\tilde p_{\text{SINLA}}(x_i \, | \, \theta, y)$ as follows
$$
\tilde p(x_i \, | \, y) = \sum_z \tilde p_{\text{SINLA}}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega(z).
$$
We want to do this because the new approximation should be more accurate than taking marginals of a joint Gaussian approximation.
At the extreme, you could just compute a Laplace approximation for each marginal, integrating out all other elements of the latent field, $x_{-i}$, together with the fixed effects $\theta$.
This would be an accurate approach, but not computationally feasible for large models.
What we hope to achieve is an approximation which gets to most of the accuracy benefits of the Laplace approximation approach, but without the high computational cost.

@@

The Gaussian approximation $p_\text{G}(x \, | \, \theta, y)$ is previously used in computing the Laplace approximation to $p(\theta \, | \, y)$
$$
\begin{equation}
\tilde p(\theta \, | \, y) \propto \frac{p(x, \theta, y)}{\tilde p_\text{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat x(\theta)} = \frac{p(y, \hat x(\theta), \theta)}{\det \left(\frac{1}{2 \pi} H(\theta) \right)^{1/2}}, (\#eq:pthetagiveny)
\end{equation}
$$
such that one easy way to calculate it is to set `random = "x"` in the `TMB` template to integrate the latent field out, then recover the Gaussian approximation from there, as we have done in the function `h` above.
Let's optimise `h` to obtain estimates $\hat \theta$ which maximise $\tilde p(\theta \, | \, y)$:

@@

We can also look at the parameters -- that is, mode vector and precision matrix -- of the Gaussian approximation at the mode `opt_theta$par` above.
Note that in `aghq`, what we would do is create a quadrature grid, then get the mode vector and precision matrix for each value of the quadrature grid over $\theta$.
By just sticking to the mode of $\theta$, we are effectively doing `aghq` with $k = 1$ in each dimension, i.e. one grid point, corresponding to empirical Bayes type of inference.

@@

x_hat <- mm[h$env$random]
x_hat <- rlang::duplicate(x_hat) #' Insurance...
length(x_hat) #' The length is 30

#' Hessian evaluated at the mode
x_H <- h$env$spHess(mm, random = TRUE)
x_H <- rlang::duplicate(x_H)
dim(x_H) #' With dimension [30 x 30]
```

@@

Given $\theta$, we compute the `mode` and `H` of $x$ (corresponding to `phi_prev`) using `h`.
We then concatenate and evaluate using `f` to get the numerator, multiplying by -1 because `f$fn` gives the negative log-likelihood, and we want the log-likelihood.
In the denominator, we use `determinant` on the log-scale as it is more numerically stable than e.g. `log(det(...))`.

```{r}
calculate_rhs <- function(theta) {
  #' Calculate mode and H for those theta
  #' TMB template remembers what you've called -- spooky
  h$fn(theta)
  mm <- h$env$last.par
  mode <- mm[h$env$random]
  H <- h$env$spHess(mm, random = TRUE)

  #' Append the calculated latent field mode
  param <- theta
  param$phi_prev <- as.numeric(mode)
  param <- param[c("beta_prev", "phi_prev", "log_sigma_phi_prev")]

@@

  #' Evaluate the log-denominator of the RHS
  denom <- 0.5 * (determinant(H, logarithm = TRUE)$modulus - (nrow(H) * log(2 * pi)))

@@

```{r}
i <- 1
mode_i <- as.numeric(x_hat[i])
var_i <- solve(x_H)[i, i]
```

@@

```{r message=FALSE, warning=FALSE}
plot_pdf2 <- plot_pdf +
  stat_function(
    data = data.frame(x = c(0.5, 3.5)), aes(x, col = "Joint Gaussian marginal"), fun = dnorm, n = 101, 
    args = list(mean = mode_i, sd = sqrt(var_i))
  ) +
  labs(x = "prev_phi[1]", y = "Posterior", col = "Approximation") +
  scale_colour_manual(values = cbpalette) +
  theme_minimal()

M <- 1000
samples_gaussian <- rnorm(M, mode_i, sqrt(var_i))

@@

For the approximation to the latent field marginal posteriors, another approach we could take is to calculate the full Laplace approximation
$$
\begin{align*}
  \tilde p(x_i \, | \, \theta, y) &\propto \frac{p(x, \theta, y)}{p_\text{G}(x_{-i} \, | \, x_i, \theta, y)} \Big\rvert_{x_{-i} = \hat x_{-i}(x_i, \theta)} (\#eq:pxigiventhetay) \\
  &= \frac{p(y, x_i, \hat x_{-i}(x_i, \theta), \theta)}{\det(\frac{1}{2 \pi} H_{-i}(x_i, \theta))^{1/2}}
\end{align*}
$$
integrating out $x_{-i}$ with a Gaussian approximation.
This is not practical in some settings as it involves recomputing the Gaussian approximation $p_\text{G}(x_{-i} \, | \, x_i, \theta, y)$ for each value of $(x_i, \theta)$ which can be too computationally expensive.
We will still go through the steps for implementing it here though for demonstration.
With of a bit of effort, it can be implemented in `TMB` by passing in data where elements $i$ and $-i$ of the latent field are explicitly named (see Appendix \@ref(appendix) for the `TMB` code):

@@

The mode and precision matrix for the Gaussian approximation $\tilde p_\text{G}(x_{-i} \, | \, x_i, \theta, y)$ are:

```{r}
get_mode_and_H <- function(obj) {
  mm <- obj$env$last.par
  mode <- mm[obj$env$random]
  mode <- rlang::duplicate(mode)
  H <- obj$env$spHess(mm, random = TRUE)
  H <- rlang::duplicate(H)
  return(list(mode = mode, H = H))
}

xminusi <- get_mode_and_H(obj)
```

@@

```{r}
x_Sigma <- solve(x_H)

gaussian_df <- lapply(1:dat$n, function(i) {
  x <- seq(-5, 5, length.out = 1000)
  mode_i <- as.numeric(x_hat[i])
  sd_i <- as.numeric(sqrt(x_Sigma[i, i]))
  data.frame(index = i, x = x, pdf = dnorm(x, mean = mode_i, sd = sd_i))
}) %>%
  bind_rows()

samples_gaussian <- lapply(1:dat$n, function(i) {
  mode_i <- as.numeric(x_hat[i])
  sd_i <- as.numeric(sqrt(x_Sigma[i, i]))
  samples <- rnorm(M, mode_i, sd_i)
  data.frame("value" = samples)
}) %>%
  bind_rows(.id = "index")
```

@@

```{r}
ks_gaussian_tmbstan <- lapply(1:length(mode), function(i) {
  samples_gaussian_i <- filter(samples_gaussian, index == i)$value
  samples_tmbstan_i <- filter(samples_tmbstan, index == i)$value
  inf.utils::ks_test(samples_gaussian_i, samples_tmbstan_i)
})

ks_laplace_tmbstan <- lapply(1:length(mode), function(i) {
  samples_laplace_i <- filter(samples_laplace, index == i)$value
  samples_tmbstan_i <- filter(samples_tmbstan, index == i)$value
  inf.utils::ks_test(samples_laplace_i, samples_tmbstan_i)
})

@@

Now we want to evaluate
$$
\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega(z).
$$

@@

The weight $\omega(z)$ is:

@@

```{r}
mode_i <- as.numeric(modesandhessians$mode[[z]][i])
Sigma <- solve(modesandhessians$H[[z]])
sd_i <- as.numeric(sqrt(Sigma[i, i]))

dnorm(x_test_value, mean = mode_i, sd = sd_i, log = TRUE)
```

@@

```{r}
dnorm(x_test_value, mean = mode_i, sd = sd_i, log = TRUE) + (- h_norm(theta)) + log(modesandhessians$weights[z])
```

To obtain the unlogged version we can use:
$$
\exp \left( \log\tilde p_\text{G}(x_i \, | \, \theta(z), y) + \log \tilde p_\text{LA}(\theta(z) \, | \, y) + \log \omega(z) \right)
$$

```{r}
exp(dnorm(x_test_value, mean = mode_i, sd = sd_i, log = TRUE) + (- h_norm(theta)) + log(modesandhessians$weights[z]))
```

@@

for(z in 1:length(modesandhessians$weights)) {
  theta <- as.numeric(modesandhessians[z, thetanames])
  mode_i <- as.numeric(modesandhessians$mode[[z]][i])
  Sigma <- solve(modesandhessians$H[[z]])
  sd_i <- as.numeric(sqrt(Sigma[i, i]))
  pdf <- pdf + exp(dnorm(x_test_value, mean = mode_i, sd = sd_i, log = TRUE) + (- h_norm(theta)) + log(modesandhessians$weights[z]))
}

@@

Let's do as we have above, but with the full Laplace approximation in place of the Gaussian.
We want to evaluate
$$
\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{LA}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega(z),
$$
so much of what we computed above in Section \@ref(aghq-gaussian-marginal) is still relevant.
The only thing that has changed is that we're now using $\tilde p_\text{LA}(x_i \, | \, \theta(z), y)$ rather than $\tilde p_\text{G}(x_i \, | \, \theta(z), y)$.

@@

As before, the weight $\omega(z)$ is:

@@

<!-- In Section \@ref(full-laplace) we directly calculate the Laplace approximation to $\int p(x, y, \theta) \text{d} \theta_{-i}$. -->
<!-- Suppose that -->
<!-- $$ -->
<!-- x \sim \mathcal{N}(\hat x(\theta), H(\theta)^{-1}) -->
<!-- $$ -->
<!-- then by the conditional distributions of a multivariate Gaussian distribution $x_{-i} \, | \, x_i$ is Gaussian with mode -->
<!-- $$ -->
<!-- \hat x(\theta)_{-i} + \Sigma(\theta)_{-i, i} \Sigma(\theta)^{-1}_{i, i} (x_i - \hat x (\theta)_i) (\#eq:conditionalmode) -->
<!-- $$ -->
<!-- and precision matrix -->
<!-- $$ -->
<!-- H(\theta)_{-i, -i} (\#eq:conditionalprecision) -->
<!-- $$ -->

<!-- We follow Wood (2019) Section 2. -->
<!-- In this section we omit dependence on $\theta$ such that: $\hat x = \hat x(\theta)$, $H = H(\theta)$, $\hat x_{-i}(x_i) = \hat x_{-i}(x_i, \theta)$ and $H(x_i) = H_{-i, -i}(x_i, \theta)$. -->

@@

<!-- The first task is to "compute the Cholesky factor of $H_0 := H_{-i, -i}$ by update of the Cholesky factor of $H$". -->

@@

<!-- mm <- h$env$last.par -->
<!-- mode <- mm[h$env$random] -->
<!-- H <- h$env$spHess(mm, random = TRUE) -->
<!-- H0 <- H[-i, -i]  -->

<!-- L0 <- chol(H0) -->
<!-- ``` -->

<!-- We have that $H_0 = L_0 {L_0}^\top$: -->

<!-- ```{r} -->
<!-- sum(abs((L0 %*% t(L0)) - H0)) < 10E-9 -->
<!-- ``` -->

@@

<!-- Use Newton's method with fixed Hessian $H_{-i, -i}$ to find $\hat x_{-i}(x_i)$ starting from \@ref(eq:conditionalmode). -->
<!-- We start by computing the conditional mode for the first grid point: -->

<!-- ```{r} -->
<!-- Sigma <- solve(H) -->
<!-- conditional_mode <- mode[-i] + Sigma[-i, i] * Sigma[i, i]^{-1} * (x_grid[1] - mode[i]) -->
<!-- ``` -->

<!-- Note that for this model the latent field is IID, so the conditional mode is the same as the mode. -->
<!-- Now we use Newton's method to find $\hat x_{-i}(x_i)$. -->
<!-- This requires some function $f(\hat x_{-i}(x_i), \ldots) = 0$. -->

@@

<!-- For $j = 1, \ldots, J - 1$ compute -->
<!-- $$ -->
<!-- q = H_0 \Delta_j + u_j \mathcal{D}_{2j}u_j^\top \Delta_j -->
<!-- $$ -->
<!-- and -->
<!-- $$ -->
<!-- g = \Delta_x \log p(\hat x_{-i}(x_i) + \Delta_j^i, y, \theta). -->
<!-- $$ -->
<!-- Then compute the matrix -->
<!-- $$ -->
<!-- u_{j + 1} = \{q(\Delta_j^\top j)^{-1/2}, g_{-i}(\Delta_j^\top g_{-i})^{-1/2}, u_j\}. -->
<!-- $$ -->

@@

<!-- Compute the determinant approximation -->
<!-- $$ -->
<!-- \det(H_1) = \det(H_0) \det(I_{2J} + u_J^\top H_0^{-1} u_J \mathcal{D}_{2j}). -->
<!-- $$ -->

@@

| Description | Here          | Wood (2019) notation | Notes    |
|:------------|:--------------|:---------------------|:---------|
| Random effects | $x$        | $\beta$              | Stringer (2021) use $w$
| Random effects which maximise (the / some) objective for given $\theta$ | $\hat x (\theta)$ | $\hat \beta$ | Wood omits dependence on $\theta$ |
| Random effects which maximise (the / some) objective for given $\theta$, subject to $x_i$ being fixed | $(x_i, \hat x_{-i}(x_i, \theta))$ | $\tilde \beta$ | We have $\hat x_{-i}(x_i, \theta))$ having dimension $n - 1$ |
| Hessian / precision matrix of the negative log-likelihood with respect to $x$ at $\hat x(\theta)$ | $H(\theta)$ | $H$ | Unsure if it's preferable to use Hessian or precision notation here |
